Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife

### 2023-04-13 analysis/paper/paper.Rmd
writing in progress

As expected, none of the projection methods yield accurate spatial projections of road presence (i.e. the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`). Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`). All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)). The DLCP projection method had slightly lower performance than the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.

@@

- we selected revelstoke because it is topographically complex, so a good challenge. Hardy result in eastern Canadian shield showed better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for. Stream crossing behaviour in Revelstoke landscape could be improved with more a detailed high resolution cost surface. But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain. In particular, none of these simple algorithms will project switchbacks on steep slopes. A related fundamental limitation is that none of these algorithms can project tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases. 
- Aspatial overall performance metrics can be difficult to interpret. To more deeply understand variation in projection accuracy we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory result finding that me Better harvest & existing road data would allow us to more distinguish errors in projections from errors in the input data (cite lit in coops paper)
 - Outside cutblocks, we see very little variation in performance among our method variants - any of them are better than nothing. Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm. Hardy algorithm most accurately projected road density within blocks - could up density of landing pts in DCLP/MST algorithms to match this. But note that these algorithms can get the right density without replicating the pattern or mechanism. Hardy density is high in part because the algorithm is prefering turns over straigt lines straight lines, which may be an artefact of searching for nearest destinations >200 m away on a 100 m grid. Projected wiggles are quite different from observed switchbacks. 
- Focussed here on forest resource roads. Note the algorithms are also likely to work for other situations where the objectives of road building are to access target locations and minimize cost. Not a useful algorithm for projecting linear features that were constructed for some other purpose such as seismic testing (e.g. Figure \@ref(fig:ftNelson)), or for circumstances where information about target locations is not available. When considering transferring these results to other contexts, also note that in this example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. It would be interesting to test performance in a circumstance where target locations are far from existing roads. 
- If realistic switchbacks in mountainous terrain is a key goal then none of these algorithms will do. But before investing in more complexity recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For example, models of caribou movement often include distance to roads as a predictor, and at a landscape scale a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates (). In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). For wildlife response models informed by distance to road or buffered disturbance footprint, somewhat crude road projections are likely good enough. When constructing new wildlife response models intended for projection, it be helpful to select predictor variables such as distance to road, road presence within grid cells, and disturbance footprint that are more easily projectable than road density (Bodner).
- There are a bewildering array of possible metrics of connectivity and fragmentation we did not consider in this paper. In general, performance will vary among metrics, and it is difficult to tell what is good enough without reference to particular circumstances (objectives, species, landscapes). Rather than expanding the array of road performance metrics considered, we believe our efforts are better spent making the road projection tools easy to use, to enable users devise performance tests appropriate to their particular circumstances and goals. The best performance metrics will be those most closely linked to the outcome of interest. For example, where the outcome of interest is space use or connectivity projected from more (e.g. Whittington, Labadie) or less (samc, pither etc) species-specific movement models consider testing sensitivity of these models to the difference between observed and retrospectively projected roads. We hope R package and open workflow associated with this paper can help enable retrospective projection and relevant performance assessment in other contexts. However, we also recognize there is work associated with acquiring data to construct cost surfaces & retrospectively project roads. For those seeking a lower effort option, we also provide our output maps and an example script () to enable calculation of alternative performance metrics of interest for our Revelstoke example.

>>>>

As expected, none of the projection methods yield accurate spatial projections of road presence (i.e. the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`). Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`). All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)). The DLCP projection method had comparable performance to the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.

@@

- we selected revelstoke because it is topographically complex, so a good challenge. Hardy result in eastern Canadian shield showed better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for. Stream crossing behaviour in Revelstoke landscape could be improved with more a detailed high resolution cost surface. But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain. None of these simple algorithms will project switchbacks on steep slopes. A related fundamental limitation is that none of these algorithms can project tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases. 
- Aspatial overall performance metrics can be difficult to interpret. To more deeply understand variation in projection accuracy we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results. Better harvest & existing road data would allow us to more distinguish errors in projections from errors in the input data (cite lit in coops paper)
 - Outside cutblocks, we saw very little variation in performance among our method variants - any of them are somewhat better than nothing, but they don't make a big differene, partly because in this example projected new roads over fairly short distances, and new landings were always fairly close to existing roads or other landings. Would be interesting to test in a context where targets are further from existing roads, but outside cutblocks our DLCP algorithm is very similar to Hardy's algorithm, so expect similar performance. Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm. Hardy algorithm most accurately projected road density within blocks - could up density of landing pts in DCLP/MST algorithms to match this. But examination of spatial patterns clarifies that algorithms can get the right density without replicating the pattern or mechanism. Hardy density is high in part because the algorithm is prefering turns over straigt lines straight lines, which may be an artefact of searching for nearest destinations >200 m away on a 100 m grid. Projected wiggles are quite different from observed switchbacks. 
- Focussed here on forest resource roads. Note the algorithms could also work in other situations where the objectives of road building are to access target locations and minimize cost. Not a useful algorithm for projecting linear features that were constructed for some other purpose such as seismic testing (e.g. Figure \@ref(fig:ftNelson)), or for circumstances where information about target locations is not available. 
- If realistic switchbacks in mountainous terrain is a key goal then none of these algorithms will do. But before investing in more complexity recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For example, models of caribou movement often include distance to roads as a predictor, and at a landscape scale a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates (). In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). For wildlife response models informed by distance to road () or buffered disturbance footprint (), somewhat crude road projections are likely good enough. When constructing new wildlife response models intended for projection, it may be helpful to select these more easily projectable (Bodner) predictors.
- There are a bewildering array of possible metrics of connectivity and fragmentation we did not consider in this paper. In general, performance will vary among metrics, and it is difficult determine what is good enough without reference to particular circumstances (objectives, species, landscapes). Rather than expanding the array of road performance metrics considered, we believe our efforts are better spent making road projection tools easy to use, to enable users devise performance tests appropriate to their particular circumstances and goals. The best performance metrics will be those most closely linked to the outcome of interest. For example, where the outcome of interest is space use or connectivity projected from more (e.g. Whittington, Labadie) or less (samc, pither etc) species-specific movement models consider testing sensitivity of these models to the difference between observed and retrospectively projected roads. We hope R package and open workflow associated with this paper can help enable retrospective projection and relevant performance assessment in other contexts. However, we also recognize there is work associated with acquiring data to construct cost surfaces & retrospectively project roads. For those seeking a lower effort option, we provide our output maps and an example script () to enable calculation of alternative performance metrics of interest for our Revelstoke example.
